---
title: "HW3"
author: "Sinclair Stockard"
date: "October 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1a

There are 28,947 probes.

```{r}
setwd('/Users/sincl/OneDrive/Desktop')

load(file.path("ProstateData", "expression.data.rda"))
load(file.path("ProstateData", "clinical.info.rda"))
load(file.path("ProstateData", "gene.info.rda"))

dim(expression.data)

expression.data[1:5,1:5]

```

## Question 1b

There are 10,204 unique genes.

```{r}

str(gene.info)

```

## Question 1c

There are 12,248 unassociated probes.

```{r}

summary(gene.info)

```

## Question 1d

There are 112 samples.

```{r}

dim(clinical.info)

clinical.info[1:5,1:5]

```

## Question 1e

41 samples normal (Status = N)
62 samples tumor (Status = T)
9 samples lymphnode metastases (Status = L)

```{r}

str(clinical.info)

Status <- clinical.info$Status

freqtable <- table(Status)
t = as.data.frame(freqtable)

t

```
## Question 2a

```{r}
T <- (which(clinical.info$Status == "T"))
L <- (which(clinical.info$Status == "L"))
filtdata <- expression.data[, c(T, L)]
filtclin <- clinical.info[c(T, L),]
filtclin$Status <- factor(filtclin$Status)

```
## Question 2b

```{r}

library(ClassComparison)
mtt <- MultiTtest(filtdata, filtclin$Status)
hist(mtt@t.statistics, breaks=100, col='skyblue')

```

## Question 2c

```{r}

hist(mtt@p.values, breaks=100, col='pink')

```
## Question 2d

There are 5358 p-values < .01. This is more than what is expected by chance (0.01 * 28947 = 289.5).

```{r}

p.val <- mtt@p.values
sum(p.val < 0.01)


```

## Question 2e

There are 8574 p-values < .05. This is more than what is expected (1447.35).

```{r}

sum(p.val < 0.05)

```

## Question 3a

```{r}

bum <- Bum(mtt@p.values)
hist(bum)

```

## Question 3a

There are 3666 genes differentially expressed. 

```{r}

countSignificant(bum, alpha=0.01, by="FDR")

```
## Question 3b

There are 6568 genes differentially expressed. 

```{r}

countSignificant(bum, alpha=0.05, by="FDR")

```

## Question 4a

```{r}
status <- filtclin[, "Status"]

wiltest <- MultiWilcoxonTest(filtdata, status)

hist(wiltest, breaks = 100, col='orange')
```

## Question 4b

```{r}

require(ClassComparison)
wilc <- MultiWilcoxonTest(filtdata, status)

plot(wilc, prior=c(0.30, 0.40, 0.50, .60), signif=NULL)

abline(h=0)

plot(wilc, prior=0.52 , ylim=c(0,1))
abline(h=0)

```

## Question 4c 

```{r}

cutoffSignificant(wilc, prior=0.52, signif=0.9)

```

## Question 5a

Yes, similar numbers of L samples were run on new and old chips, but far more N and T samples were run on new chips. 

```{r}

summary(clinical.info)

newchip <- clinical.info[clinical.info$ChipType %in% c('new'),]

StatusNew <- newchip$Status

freqtable2 <- table(StatusNew)
t2 = as.data.frame(freqtable2) ##Distribution of sample types for 'new' chips

t2

oldchip <- clinical.info[clinical.info$ChipType %in% c('old'),]

StatusOld <- oldchip$Status

freqtable3 <- table(StatusOld)
t3 = as.data.frame(freqtable3) ##Distribution of sample types for 'old' chips

t3

```

## Question 5b

Chip Type seems to have more of an overall impact on data. All of the old samples are clustered together in one branch whereas status varies within branches. 

```{r}

expression.data2 <- as.data.frame(t(expression.data))

rownames(expression.data2) = paste(rownames(expression.data2), as.character(clinical.info$ChipType))

pca <- prcomp(expression.data2, scale = TRUE)

clust = hclust(dist(pca$x, "euclidean"), method = "ward.D2")

plot(clust, cex = .6)

```

## Question 6a

```{r}

Histology <- as.data.frame(clinical.info$Status)
rownames(Histology) = rownames(clinical.info)
colnames(Histology) = "Status"
mlm <- MultiLinearModel(Y ~ Status, Histology, expression.data)
summary(mlm)
plot(mlm)
 
```

## Question 6b

```{r}

bummlm <- Bum(mlm@p.values)
hist(bummlm)
```

## Question 6c

8516 significant genes at 1% FDR
12409 significnant genes at 5% FDR
15140 significant genes at 10% FDR
```{r}

countSignificant(bummlm, alpha=0.01, by="FDR")

countSignificant(bummlm, alpha=0.05, by="FDR")

countSignificant(bummlm, alpha=0.10, by="FDR")

```

## Question 7a

```{r}

Components <- as.data.frame(clinical.info[,c(4,6)])
mlm2 <- MultiLinearModel(Y ~ Status + ChipType, Components, expression.data)
summary(mlm2)
plot(mlm2)

```

```{r}

anova <- anova(mlm,mlm2)

summary(anova)

```

```{r}

bummlm2 <- Bum(mlm2@p.values)

countSignificant(bummlm2, alpha=0.01, by="FDR")
```

## Question 7b

```{r}

Chip <- as.data.frame(clinical.info$ChipType)
rownames(Chip) = rownames(clinical.info)
colnames(Chip) = 'ChipType'
mlm3 <- MultiLinearModel(Y ~ ChipType, Chip, expression.data)
summary(mlm3)
plot(mlm3)
```

```{r}

bummlm3 <- Bum(mlm3@p.values)

countSignificant(bummlm3, alpha=0.01, by="FDR")

```

## Question 7c

Chip Type is the more important factor. 